---
title: "Session 3"
author: "Kristin Porter"
date: "`r Sys.Date()`"
output:
   xaringan::moon_reader:
 #   css: ["default"]
---

<!-- Add CSS in a style block -->
<style>
.two-columns {
  display: flex;
}
.column {
  flex: 1;
  padding: 10px;
}
</style>

```{r setup, include=FALSE}
library(knitr)
library(xaringan)
library(xaringanExtra)
knitr::opts_chunk$set(echo = TRUE)
```

## Navigating R Studio - Left Side

**Console**
- Primary interface for running R commands interactively
- Displays output and error messages
- Supports direct command execution and script running

**Code View Pane**
- Area for writing and editing R scripts
- Syntax highlighting and code completion features
- Multiple script tabs can be opened simultaneously


<span style="color: blue;">Exercise: Let's create an R script.</span>

---
## Navigating R Studio - Lower Right

- **Files**: Manage project files and directories

- **Plots**: Display and manage generated plots and graphics

- **Packages**: Install, update, and load R packages

- **Help**: Access R documentation and search help topics

- <span style="color: gray;">**Viewer**: Render and preview HTML content and interactive visualizations</span>

- <span style="color: gray;">**Presentation**: Display R Markdown presentations</span>


<span style="color: blue;">Exercise: Let's see if `dplyr` is installed and if not, then install it.</span>

<span style="color: blue;">Exercise: Let's pull up documentation for `dplyr`.</span>

---
## Navigating R Studio - Upper Right

- **Environment**: View and manage workspace variables and data

- **History**: Track and recall previously executed commands

- <span style="color: gray;">**Connections**: Connect to databases and manage connections</span>

- <span style="color: gray;">**Git**: Version control integration for project collaboration</span>

- <span style="color: gray;">**Tutorial**: Access interactive tutorials and learning resources</span>

---
## Set-up: Tip 1 - Don't worry about "workspace"

In R Studio, go to: 
- "Edit" then
- "Settings

1. Uncheck "Restore.Rdata into workspace at start-up"

2. For "Save workspace to .Rdata on exit" switch to "Never." 

Hit "Apply" then "OK"

---
## Set-up: Tip 2 - Create an R project

### The RStudio project file:
- Sits in root directory with extension .Rproj
- When your RStudio session is running through the .Rproj, the current working directory points to the root folder where the .Rproj is saved

### Steps:
- Create a folder called "NCSU_DSA".
- Within RStudio, to File > New Project
- Select “New Directory”
- Select “New Project”
- For “Directory name:” enter “NCSU_DSA”. 
- Select subdirectory. 
- Click “Create Project”
- Navigate to "NCSU_DSA" folder in RStudio and confirm you see “NCSU_DSA.Rproj”
- Save your script here. 

---
## Quick overview of R Markdown
- **Combines Code and Text**
  - Integrates R code with narrative text, equations, and visualizations
  - Supports reproducible research and dynamic reports
- **Flexible Output Formats**
  - Can generate HTML, PDF, Word documents, slideshows, and more
  - Easily customizable templates and themes
- **Code Chunks**
  - Embed executable R code in chunks
  - Display code output inline with text
  - Supports multiple languages (Python, SQL, etc.)
- **R Notebooks**
  - A type of R Markdown that provides a more interactive experience
- **Learning Resources**
  - Comprehensive guide and reference: [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)
  - RStudio's R Markdown resources: [R Markdown Documentation](https://rmarkdown.rstudio.com/)
  
---
## Questions? 

---
# Review of R Foundations

---
## R Objects

In R, the term "object" refers to...
- anything you can manipulate
- anything assigned to a variable
- a wide range of data structures (e.g., numbers, strings vectors, lists, matrices, data frames, and functions).

Objects have:

1.  **Attributes**: These are "metadata" about the object, like its dimensions (for matrices), names (for elements in a vector or columns in a data frame), or class (which defines how the object should behave in different contexts).

2.  **Types and Structures**: The type of an object refers to its basic data type (numeric, integer, character, etc.), while the structure refers to how these types are organized (like vectors, lists, matrices).

3.  **Classes and Methods**: Objects can have specific classes, and there are methods designed to work with objects of these classes. For instance, the **`print(x)`** function behaves differently for object `x` depending on whether `x` is a numeric vector, a data frame, or a model object.


---
## R Data Types

<div class="two-columns">
  <div class="column">
    <!-- First column with text -->
1.  **Numeric**: Default data type for numbers in R. 

2.  **Character**: Strings of text. Any data enclosed in single or double quotes is considered a character type, e.g., "Raleigh".

3.  **Logical**: Boolean values, which take values of **`TRUE`** or **`FALSE`**.

4. **Factors**: Used to store ordered or unordered categorical data efficiently by using. Different functions handle factors in different ways - e.g. many regression functions automatically create dummy variables for each level.

  </div>
  <div class="column">
    <!-- Second column with code chunk -->
```{r}
# Here we have a numerical vector
a <- c(1,2,3)
class(a)

# If we add a string
b <- c(a,"cat")
print(b)

# We can create a Boolean
a_is1 <- a==1
a_is1

# We can turn a into a factor
factor.a <- as.factor(a)
```
  </div>
</div>
```
```
```
---
## R object names

1.  **Syntax Rules**: Valid object names must start with a letter, followed by a combination of letters, numbers, the period (`.`), and the underscore (`_`). They cannot start with a number or an underscore, nor can they contain special characters like `^`, `!`, `@`, `#`, `$`, `%`, `&`, `*`, `(`, `)`, `-`, `+`, `=`, `{`, `}`, `[`, `]`, `|`, `\`, `:`, `;`, `'`, `"`, `<`, `>`, `,`, `?`, `/`, or space. Avoid `.` at end of name (reserved for other purposes).

2.  **Reserved Words**: R has a set of reserved words that cannot be used as object names. These include function names, keywords, and other built-in identifiers like `if`, `else`, `repeat`, `while`, `function`, `for`, `in`, `next`, and `break`.

3.  **Case Sensitivity**: R is case-sensitive, which means that `Object`, `object`, and `OBJECT` would be considered different objects.
 and avoiding the use of periods at the end, as they are conventionally used for other purposes. 

4.  **Global Environment**: Object names in the global environment should not conflict with object names in attached packages. If a name conflict occurs, R will use the object from the most recently attached package or environment.

---
## R data structures: VECTORS

Central to R's flexibility for data analysis are its diverse data structures, which enable users to handle and manipulate data in a variety of ways. Primary data structures used in R:

1.  **Vectors**: The simplest and most common data structure in R, vectors are one-dimensional arrays that hold elements of the same type. They can be of various types themselves, such as numeric, integer, complex, logical, character, or raw.

```{r vectorEx}
# The object my.numeric.vector is a numeric vector:
my.vector <- c(1,2,3,4,5) # note that c() combines elements into a vector

# The object my.char.vector is a character vector:
my.char.vector <- c("walk","bus","bike","drive")

# The object my.logical.vector is a logical vector
my.logical.vector <- c(TRUE,FALSE,TRUE)
```

---
## R data structures: MATRICES

2.  **Matrices**: A matrix is a 2-dimensional data structure with rows and columns, where every element is of the **same data type**. It can be considered as a collection of vectors of equal length in a rectangular layout. 

```{r matrixEx}
# my.matrix is a numeric matrix, which I create by combining two vectors. 
# Each vector is a column of the matrix. 
# I combine the vectors using function cbind(). 
a <- c(1,2,3)
b <- c(4,5,6)
my.matrix <- cbind(a,b)
print(my.matrix)

# We can look at the dimensions of the matrix with the function dim(). This will return the number of rows, followed by the number of columns. 
dim(my.matrix)

# We can zoom in on sub-parts of the matrix by putting square brackets[] after a matrix object name.
# entries before the comma refer to rows
my.matrix[1:2,] # the first two rows
# entries after the comma refer the columns
my.matrix[,1] # the first column
# entries before and after column narrow into cells of the matrix
my.matrix[1,1] # value in the first row and first column
```

---
## R data structures: DATA FRAMES

3.  **Data Frames**: A data frame is a two-dimensional table-like structure that can hold columns of different types (e.g., numeric, character, and logical). Each column in a data frame can be thought of as a vector, and the length of each vector (number of rows) must be the same across all columns.

```{r dataframes}
# The `datasets` package has multiple toy data sets (data frames)
# We can look at the first five rows of the data set mtcars
head(mtcars)

# We can look at the names of the columns
names(mtcars)
```

---
```{r dataframes2}
# We can quickly look at the structure of the data frame
str(mtcars)
```

---
```{r dataframes3}
# We can get a summary of all the variables
summary(mtcars)
```

---
# Evaluations in R

- **Equals** (`==`)
  - `5 == 5` returns `TRUE`
  - `5 == 4` returns `FALSE`

- **Not Equals** (`!=`)
  - `5 != 4` returns `TRUE`
  - `5 != 5` returns `FALSE`

- **Greater Than (or Equal To)** (`>`, `>=`)
  - `5 > 4` returns `TRUE`
  - `4 >= 5` returns `FALSE`

- **Less Than (or Equal To)** (`<`,`<=`)
  - `4 < 5` returns `TRUE`
  - `5 <= 4` returns `FALSE`

---
# Logical Operators

- **And** (`&`)
  - `TRUE & TRUE` returns `TRUE`
  
  - `TRUE & FALSE` returns `FALSE`

- **Or** (`|`)
  - `TRUE | FALSE` returns `TRUE`
  
  - `FALSE | FALSE` returns `FALSE`

- **Not** (`!`)
  - `!TRUE` returns `FALSE`
  
  - `!FALSE` returns `TRUE`
  
---
## Using Evaluation Operators with Vectors
c(1, 2, 3) == c(1, 2, 4)  # c(TRUE, TRUE, FALSE)

c(1, 2, 3) > 2            # c(FALSE, FALSE, TRUE)

vec <- c(10, 20, 30)
vec[vec > 15]             # c(20, 30)

b <- c(1, 2)
b %in% c                  # c(TRUE, TRUE)

## Using Evaluation Operators with Data Frames
df <- data.frame(a = c(1, 2, 3), b = c(4, 5, 6))
df$a == 2                 # c(FALSE, TRUE, FALSE)
df[df$a == 2, ]           # Returns the second row
df[df$b > 4, ]            # Returns rows where b > 4

