---
title: "Smoking Cessation Toy Data Set"
output: pdf_document
---

# Effect of a Smoking Cessation Program on Quitting Rates

**Outcome:** Smoking Cessation: Binary variable indicating whether the participant quit smoking for at least six months (0 = No, 1 = Yes).

**Covariates:**

Age: Continuous variable (years).
sex: Binary variable (0 = Female, 1 = Male).
Smoking Intensity: Continuous variable (cigarettes per day).
Nicotine Dependence: Continuous variable (score on a standardized dependence scale).
Previous Quit Attempts: Binary variable (0 = No, 1 = Yes).
Support for Quitting: Likert scale (1 = Very Low, 2 = Low, 3 = Moderate, 4 = High, 5 = Very High).
Mental Health Status: Binary variable (0 = No issues, 1 = Issues such as anxiety or depression).
Motivation to Quit: Likert scale (1 = Very Low, 2 = Low, 3 = Moderate, 4 = High, 5 = Very High).
Socioeconomic Status: Ordinal variable (1 = Low, 2 = Lower-Middle, 3 = Middle, 4 = Upper-Middle, 5 = High).
Hospital: Categorical variable representing different hospitals where the randomization is blocked.

```{r simulate, echo=FALSE}
# Define the baseline covariates
def <- defData(varname = "age", dist = "normal", formula = 45, variance = 12)
def <- defData(def, varname = "sex", dist = "binary", formula = 0.5)
def <- defData(def, varname = "smoking_intensity", dist = "normal", formula = 20, variance = 5)
def <- defData(def, varname = "previous_quit_attempts", dist = "binary", formula = 0.4)
def <- defData(def, varname = "support_for_quitting", dist = "uniform", formula = "1;5")
def <- defData(def, varname = "mental_health_status", dist = "binary", formula = 0.3)
def <- defData(def, varname = "motivation_to_quit", dist = "uniform", formula = "1;5")
def <- defData(def, varname = "socioeconomic_status", dist = "uniformInt", formula = "1;5")
def <- defData(def, varname = "hospital", dist = "categorical", formula = "0.25;0.25;0.25;0.25")

# Add FTND score based on a normal distribution as a proxy
def <- defData(def, varname = "nicotine_dependence", dist = "normal", formula = 5, variance = 2)

# Define the treatment group (randomized within hospitals)
def <- defData(def, varname = "treatment", dist = "binary", formula = 0.5, id = "hospital")

# Define the binary outcome with a treatment effect
def <- defData(def, varname = "smoking_cessation_6months", dist = "binary",
               formula = "-2 + 0.05*age + 0.4*sex - 0.2*smoking_intensity - 0.2*nicotine_dependence + 0.3*previous_quit_attempts + 0.3*support_for_quitting + 0.2*motivation_to_quit + 0.1*socioeconomic_status - 0.5*mental_health_status + 1.5*treatment",
               link = "logit")

# Generate the data
set.seed(123)  
N <- 1000
data <- genData(N, def)

def2 <- defDataAdd(varname = "smoking_cessation_1year", dist = "binary",
                   formula = "-2 + 0.1*smoking_cessation_6months + 0.05*age + 0.4*sex - 0.2*smoking_intensity - 0.2*nicotine_dependence + 0.2*previous_quit_attempts + 0.3*support_for_quitting + 0.2*motivation_to_quit + 0.1*socioeconomic_status - 0.5*mental_health_status",
                   link = "logit")
data <- addColumns(def2,data)
print(head(data))
```

```{r messdata}

# CREATE SOME MISSINGNESS
data[sample(1:N,0.01*N),"smoking_cessation_1year"] <- NA
data[sample(1:N,0.02*N),"nicotine_dependence"] <- NA

# CREATE WEIRD VALUES
data[45,"age"] <- 7
data[500,"age"] <- 99
data[104,"age"] <- 11
data[300,"age"] <- 1
data[599,"age"] <- 1

data[sample(1:N,0.015*N),"support_for_quitting"] <- 0

# TURN SEX INTO CHARACTER
data <- data %>%
  mutate(sex = ifelse(sex == 1, "male", "female"))
num_to_change <- ceiling(0.10 * nrow(data))
indices_to_change <- sample(1:nrow(data), num_to_change)

# Function to capitalize first letter
capitalize_first <- function(x) {
  paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
}

# Apply the function to the selected rows
data$sex[indices_to_change] <- sapply(data$sex[indices_to_change], capitalize_first)

# View the updated data
print(head(data))
```

```{r separate}

# BREAK INTO TWO DATASETS FOR MERGING

RA_meas <- c("id","age","sex","hospital","treatment","smoking_cessation_6months","smoking_cessation_1year")
SU_meas <- c("id","nicotine_dependence","previous_quit_attempts","support_for_quitting","mental_health_status","motivation_to_quit","socioeconomic_status","hospital")

RA_data <- data %>% select(all_of(RA_meas))
SU_data <- data %>% select(all_of(SU_meas))

# CREATE SOME DUPLICATES IN SURVEY DATA
  # full duplicates
rows.full.dupIDs <- SU_data[sample(1:N, 5),]
  # just duplicate IDs 
rows.part.dupIDs <- SU_data[sample(1:N, 2),]
rows.part.dupIDs$id <- sample(rows.part.dupIDs$id,2) # permuting the id
SU_data <- rbind(SU_data,rows.full.dupIDs,rows.part.dupIDs)

# MAKE RA DATA LONG (ONE ROW EACH TIME POINT)
# Convert from wide to long format
RA_data_long <- RA_data %>%
  pivot_longer(
    cols = c(smoking_cessation_6months, smoking_cessation_1year),
    names_to = 'follow_up',
    values_to = 'smoking_cessation'
  )

# View the resulting data frame
head(RA_data_long)

RA_data<- as.tibble(RA_data_long)
SU_data <- as.tibble(SU_data)

save(RA_data,file="SmokingCessation_RA_data.Rda")
save(SU_data,file="SmokingCessation_SU_data.Rda")


```