---
title: "User-defined functions in R"
subtitle: "Session 4"
author: "Kristin Porter"
date: "June 7, 2024"
output:
   xaringan::moon_reader:
---

<!-- Add CSS in a style block -->
<style>
.two-columns {
  display: flex;
}
.column {
  flex: 1;
  padding: 10px;
}
</style>

<style>
  .centered {
    display: flex;
    justify-content: center; /* Horizontal centering */
    align-items: center;    /* Vertical centering */
    height: 100vh;         /* Full height of the viewport */
  }
</style>

```{r setup, include=FALSE}
library(knitr)
library(xaringan)
library(xaringanExtra)
knitr::opts_chunk$set(echo = TRUE)
xaringanExtra::use_panelset()
```


## SAS Macro Comparison

Like SAS macros, R functions...
- allow for code reuse, reducing redundancy
- take parameters to make the code flexible and adaptable

**Differences:**

*Text Substitution vs. Functional Programming:*
- SAS Macros: Operate primarily through text substitution, generating SAS code dynamically before execution.
- R Functions: Operate within the framework of R’s functional programming paradigm, directly manipulating data and performing calculations.

*Execution Timing:*
- SAS Macros: Code is generated during the macro processing phase and executed in the subsequent data step or procedure step.
- R Functions: Code is executed immediately when the function is called.

---
## Example SAS Macro

This very simple macro takes two arguments: `weight` and `height` in metric units. 
It calculates and returns `bmi`. 

```
%macro calculate_bmi(weight, height);
    %let bmi = %sysevalf(&weight / (&height**2));
    &bmi
%mend calculate_bmi;

/* Example usage */
%let bmi1 = %calculate_bmi(70, 1.75);
%put &bmi1;
```

---
## R function to accomplish same task

This very simple function takes two arguments: `weight` and `height` in metric units. 
It calculates and returns `bmi`. 

```{r calculate_bmi_simple}
calculate_bmi <- function(weight, height) {
    bmi <- weight / (height^2)
    return(bmi)
}

# Example usage for a single set of arguments
bmi1 <- calculate_bmi(weight = 70, height = 1.75)
bmi1 <- calculate_bmi(70,1.75)
```

Notes: 
- function name followed by "<- function()"
- arguments/parameters listed separated by "," inside "()"
- body of function encloses in "{}" 
- return statement not required but provides clarity, especially in complex functions 
- otherwise, the value of the last evaluated expression in the function body is returned by default
- to call function, pass in argument values in order or define explicitly

---
## Let's use this function with a data frame

```{r use_calculate_df, message=FALSE}
# Creating data frame
wh.dat <- data.frame(weight = c(65,70,70), height = c(1.75,1.75,1.60))
```

*Using base R*
```{r baseR_bmi}
# passing in weight and heigth column
wh.dat$bmi <- calculate_bmi(weight = wh.dat$weight, height = wh.dat$height)
wh.dat
```

---
*Using `dplyr`* 
```{r dplyr_bmi, message=FALSE, warning=FALSE}
library(`dplyr`)

# using mutate
wh.dat <- wh.dat %>% mutate(bmi = calculate_bmi(weight, height))
```

*Using `purrr`*
```{r purrr_bmi}
library(purrr)

# using map2
wh.dat$bmi <- map2_dbl(wh.dat$weight, wh.dat$height, calculate_bmi)

# if we want to use pmap, we need a list instead of a data frame
wh.dat_list <- as.list(wh.dat[,c("weight","height")])
# using pmap_dbl instead of pmap, it returns a vector instead of list
wh.dat$bmi <- pmap_dbl(wh.dat_list, calculate_bmi)
```

---
# Let's add to our function

Here we: 
1. Add an another argument, `system`. Note that we give it a default assignment. If we don't pass in anything, it will use the default. 
2. Use an `if` `else` sequence.
3. We add an error message for an unexpected value for `system`. 
```{r calculate_bmi_add}
calculate_bmi <- function(weight, height, system = "metric") {
  if (system == "metric") {
    bmi <- weight / (height^2)
  } else if (system == "US") {
    bmi <- (weight / (height^2)) * 703
  } else {
    stop("Invalid system. Use 'metric' or 'US'.")
  }
}

# Example usage for a single set of arguments
bmi2 <- calculate_bmi(70, 1.75, 'US')
#bmi2 <- calculate_bmi(70, 1.75, 'us')
```

---
## Let's add some documentation for our function

I'm using a specific format for my comments, following the framework of the `Roxygen` package. This is not required, but the style provides a good standard. 

```{r add_doc}

#' Calculate Body Mass Index (BMI)
#'
#' This function calculates BMI based on the user's weight & height. 
#'
#' @param weight Numeric: The weight of the individual in kilograms if using the metric
#' system, or in pounds if using the US system.
#' @param height Numeric: The height of the individual in meters if using the metric
#' system, or in inches if using the US system.
#' @param system Character: Specifies the measurement system used for the inputs.
#' Can be either "metric" or "US".
#'
#' @return A list containing the calculated BMI and the measurement system used.
#' @examples
#' calculate_bmi(70, 1.75, "metric")
#' calculate_bmi(155, 69, "US")
#'
calculate_bmi <- function(weight, height, system) {
  if (system == "metric") {
    bmi <- weight / (height^2)
  } else if (system == "US") {
    bmi <- (weight / (height^2)) * 703
  } else {
    stop("Invalid system. Use 'metric' or 'US'.")
  }
  
  return(list(BMI = bmi, System = system))
}
```

---
## Higher Order Functions

These are functions that take other functions as arguments or return functions as results.

Examples in Base R: apply, lapply, sapply, mapply

```{r lapply_ex}
square <- function(x) x^2
nums <- list(1, 2, 3, 4)
lapply(nums, square)
```

---
## Creating Higher-Order Functions

```{r simplehigherorder}
apply_function <- function(f, x) {
  f(x)
}

# Here we pass in the baseR function sqrt() for f
apply_function(sqrt, 16)  
```

---
## Anonymous Functions

These are functions defined inline without a name, often used as arguments to higher-order functions.

Base R example: 
```{r anonfun}
nums <- list(1, 2, 3, 4)
unlist(lapply(nums, function(x) x^2))
```

Usage in `purrr`:

```{r purrranon}
nums <- list(1, 2, 3, 4)
unlist(map(nums, ~ .x^2))
```

---
## Function Environments

Environments are collections of objects (like functions and variables). Each function has its own environment.

Objects created inside a function are local to that function. 
Objected outside functions are in the global environment. 

```{r envir}
x <- 10  # Global variable

modify_variable <- function() {
  x <- 5  # Local variable
  return(x)
}

modify_variable() 
print(x)  
```

---
### Debugging Functions
Common Debugging Tools:
print()
cat()
browser()
traceback()
debug()
trace()

Using browser() 
```{r browserEx}
calculate_bmi <- function(weight, height) {
  browser()
  bmi <- weight / (height^2)
  return(bmi)
}
```
Using debug(), works similar to browser, but is initiated outside function.
Using traceback(), call after get function error. 
